Light and growth form interact to shape stomatal ratio among British angiosperms

### 2017-10-21 ms/ms.Rnw
updated figure captions

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for \Sexpr{nrow(stomata)} species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (mostly trees). Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio as well as its use as functional trait for paleoecology and crop improvement.
\end{itemize}

@@

The leading, nonmutually exclusive hypotheses are that 1) open habitats favour amphistomy because CO$_2$ diffusion most strongly limits photosynthetic rate under high light and 2) herbaceous growth form favours amphistomy because traits that maximize photosynthetic rate are often under stronger selection in herbs. \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) first noted that amphistomy is most common in herbaceous plants from open habitats (i.e., with high light) of the British flora. These observations have been replicated in other studies \citep{Mott_etal_1982, Peat_Fitter_1994b, Jordan_etal_2014, Muir_2015} and may support physiological and ecological hypotheses that CO$_2$ most strongly limits photosynthesis in high light and/or photosynthesis contributes most to fitness in herbaceous plants. Under high light, CO$_2$ can strongly limit maximum photosynthetic rates, especially in thick leaves \citep{Jones_1985}. Hence, having stomata on both surfaces relieves this limitation by adding a second parallel pathway for CO$_2$ diffusion. Parkhurst (\citeyear{Parkhurst_1978}) argued that greater leaf thickness \textit{per se} selected for amphistomy, but there is little evidence for correlations between leaf thickness and stomatal ratio independent of light \citep{Mott_etal_1982, Gibson_1996, Muir_2015}. Amphistomy is correlated with open habitat in warm desert plants of western North America \citep{Mott_etal_1982, Gibson_1996}, among the Proteaceae \citep{Jordan_etal_2014}, and in continental European herbs \citep{Bucher_etal_2017}.

Stomatal ratio is also associated with growth form. In the British flora, \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) found that trees and shrubs are nearly always hypostomatous, whereas herbs from open habitats are amphistomatous. This pattern holds when data are averaged by family to coarsely control for phylogenetic nonindependence \citep{Peat_Fitter_1994b} or when using alternative classification schemes, such as Raunki\ae r life form \citep{Peat_Fitter_1994b}. Across plants from $\sim 90$ families worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}. These patterns are consistent with other data indicating that many herbaceous plants are under strong selection for high maximum photosynthetic rates \citep{Bazzaz_1979, Korner_etal_1989, Wullschleger_1993}.

Although previous comparative studies have tested whether open habitat and growth form influence stomatal ratio, we do not know if these effects are independent of one another. Open habitat and growth form may not be independent because open habitats generally consist of more short-statured, herbaceous plants. Some authors have attempted to disentangle light and growth form by contrasting herbs from open and understory habitats \citep{Salisbury_1927}. However, this is problematic if phylogenetic relationships are not controlled for, because shade species may share traits simply because they are more closely related to each other than they are to high light species. Finally, open habitat and growth form may also interact with one another. For example, amphistomy may only be favored when CO$_2$ strongly limits photosynthetic rate (e.g. in high light) \textit{and} photosynthetic rate strongly limits fitness (e.g. in herbs).

@@

The final question is important for addressing whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate, as both the light and growth form hypotheses assume. If evolved increases in stomatal ratio are mediated by shifting abaxial stomata to the adaxial surface, holding total stomatal density constant, then the overall increase in CO$_2$ diffusion would be small. In contrast, if amphistomy evolves by increasing adaxial stomatal density while holding abaxial density constant, then \textit{total} stomatal density must increase as well. Evolutionary coordination of amphistomy and high stomatal density would reinforce one another, increasing CO$_2$ supply to chloroplasts more than changes in either trait would in isolation.

To address these questions, I reanalyzed existing data on stomatal ratio, light habitat, and growth form in British angiosperms \citep{Salisbury_1927, Fitter_Peat_1994a, BEF} using phylogenetic comparative methods. The British angiosperm flora is well suited for these questions because this flora has been comprehensively surveyed for many ecologically important traits, meaning it is probably the least biased survey of stomatal trait variation. Salisbury's observations on stomata and ecology in the British flora have heavily influenced plant ecophysiology, but many of his and subsequent authors' analyses have significant limitations because of inadequate statistical methods. For example, few analyses until recently account for phylogenetic nonindependence \citep{Felsenstein_1985}, which can strongly influence inferences on stomatal traits and growth form \citep[this study did not consider light]{Kelly_Beerling_1995}. A species-level phylogeny of the entire British flora \citep{Lim_etal_2014} now allows for the first time rigorous analysis of evolutionary relationships among stomatal ratio, light, and growth form.

@@

I excluded C$_4$ (\Sexpr{nC4} species) and CAM (\Sexpr{nCAM} species) plants. I limited this investigation to angiosperms because only \Sexpr{nNonangio} non-angiosperms had stomata data. The final dataset contained \Sexpr{nrow(stomata)} species (Fig.~\ref{fig:phylo-lifeform}). The R code accompanying this paper documents these decisions in greater detail and citations to the relevant literature.

@@

I tested whether growth form, under either classification scheme, was associated with \el~among British angiosperms. I predicted that species with faster life histories, especially therophytes (annuals), would have greater \els~than species with slower life histories, especially phanerophytes (shrubs and trees). I first used a phylogenetic ANOVA assuming an Ornstein-Uhlenbeck process model fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}. However, this analysis indicated no phylogenetic signal in the regression (See the R code accompanying this paper for further detail). Specifically, the estimated $\alpha$ parameter was extremely high. In the Ornstein-Uhlenbeck model, $\alpha$ is proportional to the inverse of the phylogenetic half-life (i.e. phylogenetic signal). When there is no phylogenetic signal (i.e. high $\alpha$), regular and phylogenetic ANOVA converge on the same parameters estimates. Furthermore, statistical tests assuming there is phylogenetic signal when in fact none exists performs worse than nonphylogenetic tests \citep{Revell_2010}. Therefore, I used a regular ANOVA with Type-2 sum of squares.

@@

I compared phylogenetic linear models to test whether growth form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$. Unlike the analysis above, there was significant phylogenetic signal in this comparison (see R code). I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface. I fit models using \pkg{phylolm} and calculated Akaike Information Criteria (AIC). For these and subsequent analyses, I assumed an Ornstein-Uhlenbeck process model for the residuals with the root character state integrated over the stationary distribution. I used a 10$^4$ parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}.

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-lifeform.pdf}}
\caption{CAPTION} 
\label{fig:phylo-lifeform}
\end{figure}

@@

\begin{figure}[ht]
\centerline{\includegraphics{figures/figureS_compare-growth-forms.pdf}}
\caption{CAPTION}
\label{fig:compare-growth-forms}
\end{figure}

>>>>

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, growth form, and phylogenetic relationships for \Sexpr{nrow(stomata)} species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (shrubs and trees). Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
	\item Comparative analyses of British angiosperms reveal two major insights into physiological evolution. First, light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. Second, strong coevolution of adaxial stomatal density and light tolerance indicates that amphistomy is an important adaptation to optimally coordinate light acquisition with gas exchange. These results advance our understanding of why stomatal ratio evolves and its potential as a functional trait for paleoecology and crop improvement.
\end{itemize}

@@

The leading, nonmutually exclusive hypotheses are that 1) open habitats favour amphistomy because CO$_2$ diffusion most strongly limits photosynthetic rate under high light and 2) herbaceous growth form favours amphistomy because traits that maximize photosynthetic rate are often under stronger selection in herbs. \cite{Salisbury_1927} first noted that amphistomy is most common in herbaceous plants from open habitats (i.e., with high light) of the British flora. These observations have been replicated in other studies \citep{Mott_etal_1982, Peat_Fitter_1994b, Jordan_etal_2014, Muir_2015} and may support physiological and ecological hypotheses that CO$_2$ most strongly limits photosynthesis in high light and/or photosynthesis contributes most to fitness in herbaceous plants. Under high light, CO$_2$ can strongly limit maximum photosynthetic rates, especially in thick leaves \citep{Jones_1985}. Hence, having stomata on both surfaces relieves this limitation by adding a second parallel pathway for CO$_2$ diffusion. \cite{Parkhurst_1978} argued that greater leaf thickness \textit{per se} selected for amphistomy, but there is little evidence for correlations between leaf thickness and stomatal ratio independent of light \citep{Mott_etal_1982, Gibson_1996, Muir_2015}. Amphistomy is correlated with open habitat in warm desert plants of western North America \citep{Mott_etal_1982, Gibson_1996}, among the Proteaceae \citep{Jordan_etal_2014}, and in continental European herbs \citep{Bucher_etal_2017}.

Stomatal ratio is also associated with growth form. In the British flora, \cite{Salisbury_1927} found that trees and shrubs are nearly always hypostomatous, whereas herbs from open habitats are amphistomatous. This pattern holds when data are averaged by family to coarsely control for phylogenetic nonindependence \citep{Peat_Fitter_1994b} or when using alternative classification schemes, such as Raunki\ae r life form \citep{Peat_Fitter_1994b}. Across plants from $\sim 90$ families worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}. These patterns are consistent with other data indicating that many herbaceous plants are under strong selection for high maximum photosynthetic rates \citep{Bazzaz_1979, Korner_etal_1989, Wullschleger_1993}.

Although previous comparative studies have tested whether open habitat and growth form influence stomatal ratio, we do not know if these effects are independent of one another. Open habitat and growth form may be confounded because open habitats generally consist of more short-statured, herbaceous plants. Some authors have attempted to disentangle light and growth form by contrasting herbs from open and understory habitats \citep{Salisbury_1927}. However, this is problematic if phylogenetic relationships are not controlled for, because shade species may share traits simply because they are more closely related to each other than they are to high light species. Finally, open habitat and growth form may also interact with one another. For example, amphistomy may only be favored when CO$_2$ strongly limits photosynthetic rate (e.g. in high light) \textit{and} photosynthetic rate strongly limits fitness (e.g. in herbs).

@@

% MARKER (still revising, adding predictions here): Specifically, I predicted that species with faster life histories, especially therophytes (annuals), would have greater \els~than species with slower life histories, especially phanerophytes (shrubs and trees).

The final question is important for addressing whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate, as both the light and growth form hypotheses assume. If evolved increases in stomatal ratio are mediated by shifting abaxial stomata to the adaxial surface, holding total stomatal density constant, then the overall increase in CO$_2$ diffusion would be small. In contrast, if amphistomy evolves by increasing adaxial stomatal density while holding abaxial density constant, then \textit{total} stomatal density must increase as well. Evolutionary coordination of amphistomy and high stomatal density would reinforce one another, increasing CO$_2$ supply to chloroplasts more than changes in either trait would in isolation. Understanding selection on coordinated traits can explain the evolution of major functional trait axes and syndromes.

To address these questions, I reanalyzed existing data on stomatal ratio, light habitat, and growth form in British angiosperms \citep{Salisbury_1927, Fitter_Peat_1994a, BEF} using phylogenetic comparative methods. The British angiosperm flora is well suited for these questions because this flora has been comprehensively surveyed for many ecologically important traits, meaning it is probably the least biased survey of stomatal trait variation. Salisbury's observations on stomata and ecology in the British flora have heavily influenced plant ecophysiology, but many of his and subsequent authors' analyses have significant limitations because of inadequate statistical methods. For example, few analyses until recently account for phylogenetic nonindependence \citep{Felsenstein_1985}, which can strongly influence inferences on stomatal traits and growth form \citep[this study did not consider light]{Kelly_Beerling_1995}. A species-level phylogeny of the entire British flora \citep{Lim_etal_2014} now allows for the first time a rigorous analysis of evolutionary relationships among stomatal ratio, light, and growth form.

@@

I excluded C$_4$ (\Sexpr{nC4} species) and CAM (\Sexpr{nCAM} species) plants. I limited this investigation to angiosperms because only \Sexpr{nNonangio} non-angiosperms had stomata data. The final dataset contained \Sexpr{nrow(stomata)} species (Fig.~\ref{fig:phylo-lifeform}, \ref{fig:phylo-habit}). The R code accompanying this paper documents these decisions in greater detail and citations to the relevant literature.

@@

I tested whether growth form, under either classification scheme, was associated with \el~among British angiosperms. I first used a phylogenetic ANOVA assuming an Ornstein-Uhlenbeck process model fit using \pkg{phylolm} version 2.5 \citep{Ho_Ane_2014}. However, this analysis indicated no phylogenetic signal in the regression (See the R code accompanying this paper for further detail). Specifically, the estimated $\alpha$ parameter was extremely high. In the Ornstein-Uhlenbeck model, $\alpha$ is proportional to the inverse of the phylogenetic half-life (i.e. phylogenetic signal). When there is no phylogenetic signal (i.e. high $\alpha$), regular and phylogenetic ANOVA converge on the same parameters estimates. Furthermore, statistical tests assuming there is phylogenetic signal when in fact none exists performs worse than nonphylogenetic tests \citep{Revell_2010}. Therefore, I used a regular ANOVA with Type-2 sum of squares.

@@

I compared phylogenetic linear models to test whether growth form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$. Unlike the analysis above, there was significant phylogenetic signal in this comparison (see R code). I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface. I fit models using \pkg{phylolm} and calculated Akaike Information Criteria (AIC). For these and subsequent analyses, I assumed an Ornstein-Uhlenbeck process model for the residuals with the root character state integrated over the stationary distribution. I used 10$^4$ parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}.

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-lifeform.pdf}}
\caption{Phylogenetic diversification of stomatal ratio follows growth form and light tolerance. At the center is the phylogenetic tree for \Sexpr{nrow(stomata)} species of British angiosperms. For each species, the green wedges indicate Raunki\ae r life form and the orange wedges indicate \el. The outer circle indicates the stomatal ratio for each species. As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ration means that most stomata are on the lower surface.} 
\label{fig:phylo-lifeform}
\end{figure}

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-habit.pdf}}
\caption{Phylogenetic diversification of stomatal ratio follows growth form and light tolerance. At the center is the phylogenetic tree for \Sexpr{nrow(stomata)} species of British angiosperms. For each species, the green wedges indicate plant habit and the orange wedges indicate \el. The outer circle indicates the stomatal ratio for each species. As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ration means that most stomata are on the lower surface.} 
\label{fig:phylo-habit}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics{figures/figureS_compare-growth-forms.pdf}}
\caption{Raunki\ae r lifeform and plant habit broadly overlap. The dot chart shows for each Raunki\ae r lifeform, the proportion that overlap with a given plant habit. For example, phanerophytes are mostly trees and shrubs, geophytes are all perennial, therophytes are mostly annuals, and so forth.}
\label{fig:compare-growth-forms}
\end{figure}

### 2017-10-21 ms/ms.tex
updated figure captions

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for 372 species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (mostly trees). Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio as well as its use as functional trait for paleoecology and crop improvement.
\end{itemize}

@@

I excluded C$_4$ (3 species) and CAM (2 species) plants. I limited this investigation to angiosperms because only 4 non-angiosperms had stomata data. The final dataset contained 372 species (Fig.~\ref{fig:phylo-lifeform}). The R code accompanying this paper documents these decisions in greater detail and citations to the relevant literature.

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-lifeform.pdf}}
\caption{CAPTION} 
\label{fig:phylo-lifeform}
\end{figure}

@@

\begin{figure}[ht]
\centerline{\includegraphics{figures/figureS_compare-growth-forms.pdf}}
\caption{CAPTION}
\label{fig:compare-growth-forms}
\end{figure}

>>>>

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, growth form, and phylogenetic relationships for 372 species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (shrubs and trees). Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
	\item Comparative analyses of British angiosperms reveal two major insights into physiological evolution. First, light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. Second, strong coevolution of adaxial stomatal density and light tolerance indicates that amphistomy is an important adaptation to optimally coordinate light acquisition with gas exchange. These results advance our understanding of why stomatal ratio evolves and its potential as a functional trait for paleoecology and crop improvement.
\end{itemize}

@@

I excluded C$_4$ (3 species) and CAM (2 species) plants. I limited this investigation to angiosperms because only 4 non-angiosperms had stomata data. The final dataset contained 372 species (Fig.~\ref{fig:phylo-lifeform}, \ref{fig:phylo-habit}). The R code accompanying this paper documents these decisions in greater detail and citations to the relevant literature.

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-lifeform.pdf}}
\caption{Phylogenetic diversification of stomatal ratio follows growth form and light tolerance. At the center is the phylogenetic tree for 372 species of British angiosperms. For each species, the green wedges indicate Raunki\ae r life form and the orange wedges indicate \el. The outer circle indicates the stomatal ratio for each species. As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ration means that most stomata are on the lower surface.} 
\label{fig:phylo-lifeform}
\end{figure}

@@

\begin{figure}[ht]
\centerline{\includegraphics[width=\textwidth]{figures/figure_phylo-habit.pdf}}
\caption{Phylogenetic diversification of stomatal ratio follows growth form and light tolerance. At the center is the phylogenetic tree for 372 species of British angiosperms. For each species, the green wedges indicate plant habit and the orange wedges indicate \el. The outer circle indicates the stomatal ratio for each species. As shown in the legend above, greater stomatal ratio means stomata are more evenly distributed across both leaf surfaces; lower stomatal ration means that most stomata are on the lower surface.} 
\label{fig:phylo-habit}
\end{figure}

\begin{figure}[ht]
\centerline{\includegraphics{figures/figureS_compare-growth-forms.pdf}}
\caption{Raunki\ae r lifeform and plant habit broadly overlap. The dot chart shows for each Raunki\ae r lifeform, the proportion that overlap with a given plant habit. For example, phanerophytes are mostly trees and shrubs, geophytes are all perennial, therophytes are mostly annuals, and so forth.}
\label{fig:compare-growth-forms}
\end{figure}
